Complex and informative dependencies propagate throughout ecology and can be revealed using copulas

### 2019-05-14 Paper.Rmd
finally got the length down to journal limits again

@@

@@

\noindent \emph{The number of words in the abstract:}

\noindent \emph{The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends):}

@@

\noindent \emph{The number of references:}

@@

All branches of ecology study relationships among environmental and biological variables. However, ubiquitously used approaches to studying such relationships, based on correlation and regression, provide only a small slice of the complex information contained in the relationships. Other statistical tools exist that provide a complete description of relationships between variables, based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, but not in ecology. We here explore the concepts that underpin copulas and examine the potential for those concepts to improve our understanding of ecology. We find that informative copula structure in dependences between variables is exceedingly common across all the environmental, species-trait, phenological, population, community, and ecosystem functioning datasets we considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more strongly related in their left compared to right tails, or vice versa. Copula structure and tail dependence can help reveal causal relationships between variables. We describe mechanisms by which observed copula structure and tail dependence can arise in ecological data, including a Moran-like effect whereby dependence structures between environmental variables are inherited by ecological variables; and asymmetric or nonlinear influences of environments on ecological variables, such as under Liebig's law of the minimum. We describe consequences of copula structure, including impacts on extinction risk and the stability through time of ecosystem services. By documenting the importance of a complete description of dependence, advancing conceptual frameworks, and demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

@@

We introduce the copula families we used, and 
measures of tail dependence. 
Conceptually, a bivariate copula is a pdf
on the unit square with uniform maginals. 
The copula is actually defined instead as the corresponding 
cummulative dstribution function (cdf), but both functions contain the same information.
If $(X,Y)$ is a continuous random
variable and the cdfs of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in $[0,1] \times [0,1]$, and has 
uniform marginal distributions. 
The normal copula family (henceforth abreviated N) is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a variety of 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R language. Visualizations of the copulas we used are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence and the `VineCopula` documentation
for further background.

We tested the observational hypothesis (O in Fig. \ref{MasterFigure}) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using the `BiCopEst` function in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The method performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part of the observational hypothesis
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we carried out the bivariate analysis
for all possible pairwise combinations of locations, and compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of pairs showing stronger lower-tail dependence out of 
the total number of non-independent location pairs considered.

In addition to our model selection approach, we also used nonparametric approaches to 
provide greater 
confidence in our tests of the observational hypothesis. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All the variables we study were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association
between the bounds.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
for points between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

For large datasets (large $n$),
we used $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we considered 
multiple bands $(l_b,u_b)$ to understand how dependence
varies in different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$,
abbreviating $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"). Likewise $\Ps_l=\Ps_{0,.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, and $\Dsq_u=\Dsq_{0.5,1}$.

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "\\textit{Ceratium furca} abundance","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common, we addressed the causal hypotheses (Fig. \ref{MasterFigure}, A-D)
using models. We did not systematically 
test the first two parts of causal hypothesis A (Fig. \ref{MasterFigure})
because nonlinear and thresold effects of 
environmental variables on ecological variables are already well documented in
the literature (however, see Discussion material about the green spruce aphid). To test 
the third part of causal hypothesis A, we formulated a model in which 
an environmental variable influences an ecological variable
more strongly in one of its tails, and explored whether the ecological variable
could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below (causal hypotheis B on Fig. \ref{MasterFigure}) 
does not operate here.

@@

Causal hypothesis B is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
Let $E_i(t)$ again be the ecological variable, $i=1,2$.
We initially assume the AR(1) dynamics 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The $\epsilon_i(t)$ represent environmental noise and are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is very general, and could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane or CO$_2$, or other quantities.

To explore the importance of nonlinear dynamics for causal hypothesis B, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the three model setups above, $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If causal hypothesis B is
correct, plots should be similar for populations and noises.

Causal hypothesis C was not tested. Our next 
causal hypothesis, D on Fig. \ref{MasterFigure}, 
is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This proposed mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in observed character values
across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme left-tail dependence, 2) moderate left-tail dependence,
3) symmetric tail dependence, 4) moderate right-tail dependence, and 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copula structures, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
could produce outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$ used,
thus confirming the third part of causal hypothesis A for the model we used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry of tail dependence was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) spatial dependence in 
the noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-N copula structure, but did not always reflect the 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the most extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}).

@@

Causal hypothesis B (Fig. \ref{MasterFigure}), i.e. that copula structure 
in the spatial dependence of ecological variables can arise through
Moran effects, was largely confirmed; however, similarities between environmental- and 
ecological-variable copula structure were reduced when dynamics were strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.

>>>>

Copulas and their importance in ecology

@@

The importance of copulas

The importance of copulas in ecology

@@

\noindent \emph{The number of words in the abstract:} 250

\noindent \emph{The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends):} 7493

@@

\noindent \emph{The number of references:} 31

@@

All branches of ecology study relationships among environmental and biological variables. However, 
ubiquitously used approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or vice versa. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also reveal causal relationships between variables. We describe consequences of copula structure, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

@@

Conceptually, a bivariate copula is a random variable 
on the unit square, $[0,1] \times [0,1]$, with uniform maginals. 
If $(X,Y)$ is a continuous random
variable and the cummulative distribution functions of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$,
which takes values in $[0,1] \times [0,1]$ and has 
uniform marginals. 
We introduce the copula families we used, and 
measures of tail dependence.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

We tested the observational hypothesis (Fig. \ref{MasterFigure}, O) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part 
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we carried out the bivariate analysis
for all possible pairwise combinations of locations, and 
compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was 
greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of non-independent 
location pairs showing stronger lower-tail dependence.

In addition to our model selection approach, we also used nonparametric approaches to 
provide greater 
confidence in our tests of the observational hypothesis. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

For large datasets (large $n$),
we used $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we considered 
multiple bands $(l_b,u_b)$ to understand how dependence
varies in different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$,
abbreviating $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"). Likewise $\Ps_l=\Ps_{0,.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, $\Dsq_u=\Dsq_{0.5,1}$.

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results of testing the observational hypothesis for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "\\textit{Ceratium furca} abundance","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common, we addressed the causal hypotheses (Fig. \ref{MasterFigure}, A-D)
using models. We did not systematically 
test the first two parts of A 
because nonlinear and thresold effects of 
environmental variables on ecological variables are already well documented 
(however, see Discussion material about the green spruce aphid). To test 
the third part of hypothesis A, we formulated a model in which 
an environmental variable influences an ecological variable
more strongly in one of its tails, and explored whether the ecological variable
could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below (B on Fig. \ref{MasterFigure}) 
does not operate here.

@@

Hypothesis B is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
Let $E_i(t)$ again be an ecological variable, $i=1,2$.
We initially assume the AR(1) dynamics 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is very general, and could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.

To explore the importance of nonlinear dynamics for hypothesis B, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when it is strong, it brings the nonlinearities of the model into play.

For each of the three model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If causal hypothesis B is
correct, plots should be similar for populations and noises.

Hypothesis C was not tested. Our next 
causal hypothesis, D on Fig. \ref{MasterFigure}, 
is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in observed character values
across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Our model with asymmetric environmental effects 
could produce outputs with visually apparent asymmetry 
of tail dependence, to an extent that depended on the value of $b$,
thus confirming the third part of hypothesis A for our model. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetry of tail dependence was strong; for 
$b=\pm 0.5$, asymmetry was weaker but still apparent
(Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) spatial dependence in 
effects of noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) spatial dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
ecological outputs had non-N copula structure, but did not always reflect 
asymmetry of tail dependence, perhaps because those methods measured
dependence in the extreme tails. Results using our nonparametric statistics, however,
reflected the visually apparent asymmetry (Tables \ref{SM-tab_stat_npa_asym}).

@@

Causal hypothesis B (Fig. \ref{MasterFigure}) was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when dynamics were strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
